---
title: "Minerva Bucketing: Euclidean & Agnes"
output:
  html_notebook:
    toc: yes
  html_document:
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
---
```{r setup, include=TRUE, message=FALSE}
library(tidyverse)        #for data management, manipulation, and visualization
library(cluster)          #for PAM and Agnes clustering
library(NbClust)          #for cluster count horserace
library(janitor)          #for easy frequency tables with tabyl()
library(factoextra)       #for cluster evaluation, distance, and PCA plots
library(openxlsx)         #for write.xlsx()
library(fmsb)             #for radar plots
```

Read in PITF data and IV Variables used in paper: 
  
```{r paper_vars, include=TRUE}
df <- readxl::read_xlsx('data/PRIO_2yrpeace_2yrlag.xlsx')
df <- df %>% relocate(v2x_libdem, .after= polity_squared) %>% select(-polity_squared ) %>% rename(polity_squared = v2x_libdem)
set5 <- c('polity_squared', 
          'PopYouthBulgeBy15_new', 
          'gdppc_ln',
          'nAC_perc',
          'imr_annual_mean_centered',
          'sum_PDIS_EDIS')

paper_vars_preferred <- set5
```

Scale all IVs that will be used in cluster analysis.  
Filter the dataframe to only the observations that are complete.  
Move the `country_year` variable to the index for better visualizations.  
standardization used in agnes/daisy for numeric variables is using mean abs deviation


```{r make_cluster_data, include=TRUE}
cluster_data <- df %>% 
  select(c(country,year,country_year_onset,country_year_iv,paper_vars_preferred)) %>% 
  filter(complete.cases(.)) %>%
  #mutate_at(vars(polity_squared:sum_PDIS_EDIS), 
  #          list(scale = ~as.numeric(scale(., center=TRUE, scale=meanabsdev(.))))) %>% 
  #mutate_at(vars(polity_squared:sum_PDIS_EDIS), 
  #          list(scale = ~as.numeric(scale(.)))) %>% 
  mutate_at(vars(polity_squared:sum_PDIS_EDIS), 
            list(scale = ~as.numeric(scale(., center=min(.), scale=max(.)-min(.))))) %>% 
  rowid_to_column(.) %>% 
  column_to_rownames('country_year_onset')

```

Now, the dataframe is prepared for generating a dissimilarity matrix. 

## Notes on distance metric used  

The manuscript presents the use of Euclidean distance for the primary distance metric. The `daisy()` function from the `cluster` package allows for all alternatives tested in the sensitivity analysis with the `metric =` argument. 

```{r dist_matrix, include=TRUE}
dist_matrix <- 
  cluster::daisy(
    cluster_data %>%
      select(polity_squared_scale:sum_PDIS_EDIS_scale),
    stand = FALSE,
    metric = 'euclidean')
```

```{r dist_matrix_plot, include=TRUE, fig.height=10}
(dist_matrix_plot <- fviz_dist(dist_matrix, 
                               gradient = list(low = "red", 
                                               mid = "white", 
                                               high = "blue")) + 
    ggtitle('Euclidean',
            subtitle = 'Civil War'))

```

## Linkage method of clustering  

```{r nbclust, include=TRUE, fig.height=6}
nbclust_result <- cluster_data %>% 
  select(polity_squared_scale:sum_PDIS_EDIS_scale) %>% 
  NbClust::NbClust(data = ., 
                   diss = dist_matrix,
                   distance = NULL,
                   min.nc = 3,
                   max.nc = 7,
                   method = 'ward.D2',
                   index = 'alllong')

(horse_race_viz <- fviz_nbclust(nbclust_result) + 
    labs(caption = 'Euclidean\nCivil Wars'))
(elbow_plot <- fviz_nbclust(as.matrix(dist_matrix), FUNcluster = hcut, method = "wss" )) # 'silhouette' or "wss"
```


## Agglomerative Hierarchical Clustering (Agnes)

Very Important!!!!!!  
Note that agnes(*, method="ward") corresponds to hclust(*, "ward.D2").
Ward1 (ward.D) can also be used, if the Euclidean distances (from dist()) are squared before inputing them to the hclust() using the ward.D as the method.

```{r make_agnes, include=TRUE}
set.seed(1234)
cluster_agnes<- cluster::agnes(
  dist_matrix,
  diss = TRUE,
  #metric = 'euclidean',
  #cluster_data %>% select(polity_squared_scale:sum_PDIS_EDIS_scale), 
  #stand = FALSE,
  method = "ward"
)

(dendro_viz <- fviz_dend(cluster_agnes, 
                         k=3,
                         palette='jco',
                         repel=TRUE, 
                         cex=.4) + 
    ggtitle('Agnes - Euclidean distance'))

grp <- cutree(cluster_agnes, k = 3)
options(ggrepel.max.overlaps = Inf)
(pca_viz <- fviz_cluster(list(data =     cluster_data %>%
      select(polity_squared_scale:sum_PDIS_EDIS_scale), cluster = grp),
                  palette = 'jco',
                  stand = FALSE,
                  ellipse.type = "convex", # Concentration ellipse
                  repel = TRUE, # Avoid label overplotting (slow)
                  show.clust.cent = FALSE, ggtheme = theme_minimal()) + 
    ggtitle('Agnes - Euclidean distance'))
```

```{r cluster_agnes_groups_summary, include=TRUE}
cluster_agnes_groups <- 
  cutree(as.hclust(cluster_agnes), k=3) %>% 
  as.data.frame(.) %>% 
  rowid_to_column(.) %>% 
  rename('agnes_cluster' = '.')

cluster_agnes_groups_summary <- left_join(x=cluster_agnes_groups,
                                             y=cluster_data,
                                             by='rowid')

cluster_agnes_groups_summary %>% 
  select(2:4) -> agnes_set_5

cluster_agnes_groups_summary_tabyl <- cluster_agnes_groups_summary %>%
  tabyl(agnes_cluster)

(cluster_agnes_groups_summary_tbl <- cluster_agnes_groups_summary %>% 
    group_by(agnes_cluster) %>% 
    summarise_at(vars(year, polity_squared:sum_PDIS_EDIS_scale),
                 list(median = ~median(.),
                      mean = ~mean(.),
                      sd = ~sd(.))) %>% 
    left_join(x=.,
              y=cluster_agnes_groups_summary_tabyl,
              by='agnes_cluster'))
```

## Partitioning Around Medoids (PAM) 

```{r cluster_pam_groups_summary, include=TRUE}
cluster_pam <- cluster::pam(dist_matrix, 3, diss = TRUE)

cluster_pam_groups <- as.data.frame(cluster_pam$clustering) %>% 
  rowid_to_column(.) %>% 
  rename('pam_cluster' = 'cluster_pam$clustering')

cluster_pam_groups_summary <- left_join(x=cluster_pam_groups,
                                           y=cluster_data,
                                           by='rowid')

cluster_pam_groups_summary %>% 
  select(2:4) -> PAM_set_5

cluster_pam_groups_summary_tabyl <- cluster_pam_groups_summary %>%
  tabyl(pam_cluster)

(cluster_pam_groups_summary_tbl <- cluster_pam_groups_summary %>% 
    group_by(pam_cluster) %>% 
    summarise_at(vars(year, polity_squared:sum_PDIS_EDIS_scale),
                 list(median = ~median(.),
                      mean = ~mean(.),
                      sd = ~sd(.))) %>% 
    left_join(x=.,
              y=cluster_pam_groups_summary_tabyl,
              by='pam_cluster'))
```

# Cluster membership

```{r join_agnes_pam_membership, include=TRUE}
joined_set <- left_join(x=PAM_set_5,
                        y=agnes_set_5,
                        by=c('country', 'year')) %>% 
  select(country, year, everything())
```

# Similarity analysis  

```{r most_similar, include=TRUE}
(most_similar <- as.data.frame(as.matrix(dist_matrix)) %>% 
    rownames_to_column() %>% 
    gather(country_b, similarity, 2:last_col()) %>% 
    filter(similarity != 0) %>% 
    group_by(rowname) %>% 
    slice(which.min(similarity)) %>% 
    arrange(rowname))
```

# Analysis of cluster centers  

## Agnes  

```{r agnes_center_distance, include=TRUE}
(agnes_center_distance <- cluster_agnes_groups_summary_tbl  %>% 
    mutate(country = 'Cluster center estimate') %>% 
    select(country, agnes_cluster, contains('median')) %>% 
    select(-contains('scale')) %>% 
    gather(variable, value, 3:9) %>% 
    separate(., variable, into = c('variable', 'median'), sep='_median') %>% 
    select(-median) %>% 
    spread(variable, value) %>% 
    bind_rows(x=.,
              y=cluster_data %>% 
                mutate(agnes_cluster = NA) %>% 
                select(-c(rowid, country_year_iv, polity_squared_scale:sum_PDIS_EDIS_scale))) %>% 
    rowid_to_column(.) %>% 
    mutate(country_year = ifelse(country == 'Cluster center estimate',
                                 paste(rowid, country, year, sep = '_'),
                                 paste(country, year, sep = '_'))) %>% 
    column_to_rownames('country_year'))
```

```{r dist_matrix_agnes_center, include=TRUE}
agnes_center_distance <- agnes_center_distance %>% 
  #mutate_at(vars(gdppc_ln:sum_PDIS_EDIS), 
  #          list(scale = ~as.numeric(scale(., center=TRUE, scale=meanabsdev(.))))) 
  #mutate_at(vars(gdppc_ln:sum_PDIS_EDIS), 
  #          list(scale = ~as.numeric(scale(.))))  
  mutate_at(vars(gdppc_ln:sum_PDIS_EDIS), 
            list(scale = ~as.numeric(scale(., center=min(.), scale=max(.)-min(.)))))

dist_matrix_agnes_center <- 
  cluster::daisy(
    agnes_center_distance %>%
      select(gdppc_ln_scale:sum_PDIS_EDIS_scale),
    stand = FALSE,
    metric = 'euclidean')

as.data.frame(as.matrix(dist_matrix_agnes_center)) %>% 
  rownames_to_column()
```


```{r agnes_center, include=TRUE}
center_name <- rownames(agnes_center_distance)[1:3]

(agnes_center <- as.data.frame(as.matrix(dist_matrix_agnes_center)) %>% 
    rownames_to_column() %>% 
    gather(country_b, similarity, 2:last_col()) %>% 
    filter(similarity != 0) %>% 
    group_by(rowname) %>% 
    slice(which.min(similarity)) %>% 
    arrange(rowname) %>% 
    filter(rowname %in% center_name))

(agnes_center_2 <- cluster_data %>% 
    rownames_to_column() %>% 
    inner_join(x=.,
               y=agnes_center,
               by=c('rowname' = 'country_b')))
```

```{r medoid_characteristics_agnes, include=TRUE}
(medoid_characteristics_agnes <- agnes_set_5 %>% 
    inner_join(x=.,
               y=agnes_center_2,
               by=c('country', 'year')) %>% 
    select(agnes_cluster, rowname, everything()) %>% 
    select(-c(rowname.y, similarity)) %>% 
    arrange(agnes_cluster))
```

```{r agnes_center_3, include=TRUE}
(agnes_center_3 <- agnes_set_5 %>% 
    inner_join(x=.,
               y=agnes_center_2,
               by=c('country', 'year')) %>% 
    select(rowname, agnes_cluster))
```

## PAM  

```{r pam_medoids, include=TRUE}
(pam_medoids <- cluster_pam$medoids %>% 
    as.data.frame() %>%
    rowid_to_column() %>% 
    rename('pam_cluster' = 'rowid',
           'rowname' = '.'))
```

```{r medoid_characteristics_PAM, include=TRUE}
(medoid_characteristics_PAM <- cluster_data %>% 
    rownames_to_column() %>% 
    filter(rowname %in% cluster_pam$medoids) %>% 
    left_join(x=.,
              y=pam_medoids,
              by='rowname') %>% 
    select(pam_cluster, everything()) %>% 
    arrange(pam_cluster))
```

# Most dissimilar from same cluster

## Agnes  

```{r most_dissimilar_pair_in_cluster_agnes, include=TRUE}
(most_dissimilar_pair_in_cluster_agnes <- agnes_set_5 %>% 
    mutate(rowname = paste(country, year, sep = '_')) %>% 
    select(rowname, agnes_cluster))
```

```{r most_dissimilar_same_clust_agnes, include=TRUE}
(most_dissimilar_same_clust_agnes <- as.data.frame(as.matrix(dist_matrix)) %>% 
    rownames_to_column() %>% 
    gather(country_b, similarity, 2:last_col()) %>% 
    filter(similarity != 0) %>% 
    left_join(x=.,
              y=most_dissimilar_pair_in_cluster_agnes,
              by=c('country_b' = 'rowname')) %>% 
    semi_join(x=.,
              y=agnes_center_3,
              by=c('rowname', 'agnes_cluster')) %>% 
    group_by(rowname, agnes_cluster) %>% 
    slice(which.min(-similarity)) %>% 
    arrange(agnes_cluster))
```

## PAM  

```{r most_dissimilar_pair_in_cluster, include=TRUE}
(most_dissimilar_pair_in_cluster <- PAM_set_5 %>% 
    mutate(rowname = paste(country, year, sep = '_')) %>% 
    select(rowname, pam_cluster))
```

```{r most_dissimilar_same_clust_PAM, include=TRUE}
(most_dissimilar_same_clust_PAM <- as.data.frame(as.matrix(dist_matrix)) %>% 
    rownames_to_column() %>% 
    gather(country_b, similarity, 2:ncol(.)) %>% 
    filter(similarity != 0) %>% 
    left_join(x=.,
              y=most_dissimilar_pair_in_cluster,
              by=c('country_b' = 'rowname')) %>% 
    semi_join(x=.,
              y=pam_medoids,
              by=c('rowname', 'pam_cluster')) %>% 
    group_by(rowname, pam_cluster) %>% 
    slice(which.min(-similarity)) %>% 
    arrange(rowname))
```
# Visualizing cluster meaning  

```{r build_spider_plot_data, include=TRUE}
df_cluster_summary <- cluster_agnes_groups_summary_tbl %>% 
  select(agnes_cluster, contains('mean')) %>%
  select(agnes_cluster, contains('scale')) %>% 
  select(-c(imr_annual_mean_centered_scale_median,
            imr_annual_mean_centered_scale_sd)) %>%  
  gather(variable, value, 2:7)

df_cluster_summary_2 <- df_cluster_summary %>% 
  spread(variable, value)

df_cluster_summary_minmax <- df_cluster_summary %>% 
  group_by(variable) %>% 
  summarise(min = min(value),
            max = max(value)) %>% 
  gather(trans, value, 2:3) %>% 
  spread(variable, value) %>% 
  select(-trans)

df_cluster_summary_bind <- bind_rows(df_cluster_summary_minmax, 
                                     df_cluster_summary_2) %>%
  select(-agnes_cluster) %>% 
  rename('GDP per capita' = 'gdppc_ln_scale_mean',
         'Infant Mortality Rate' = 'imr_annual_mean_centered_scale_mean',
         'Polity squared' = 'polity_squared_scale_mean',
         'Discrimination' = 'sum_PDIS_EDIS_scale_mean',
         'Border conflicts' = 'nAC_perc_scale_mean',
         'Youth bulge' = 'PopYouthBulgeBy15_new_scale_mean')
(df_cluster_summary_bind)
```

```{r global_mean_for_spiderplot, include=TRUE}
(global_mean <- cluster_data %>% 
    #select(country, year, all_of(set5)) %>% 
    #mutate_at(vars(polity_squared:sum_PDIS_EDIS), 
    #          list(scale = ~as.numeric(scale(.)))) %>% 
    group_by() %>% 
    summarise_at(vars(polity_squared_scale:sum_PDIS_EDIS_scale),
                 list(mean = ~mean(., na.rm = TRUE))) %>% 
    rename('GDP per capita' = 'gdppc_ln_scale_mean',
           'Infant Mortality Rate' = 'imr_annual_mean_centered_scale_mean',
           'Polity squared' = 'polity_squared_scale_mean',
           'Discrimination' = 'sum_PDIS_EDIS_scale_mean',
           'Border conflicts' = 'nAC_perc_scale_mean',
           'Youth bulge' = 'PopYouthBulgeBy15_new_scale_mean'))
```

```{r spider_plot_clust1, include=TRUE, fig.height=6}
spider_plot_clust1 <- radarchart(
  bind_rows(df_cluster_summary_bind[1:3,], global_mean), 
  axistype=1,
  #custom polygon
  pcol=c(rgb(0.2,0.5,0.5,0.4),
         rgb(0.5,0.5,0.5,0.3)), 
  pfcol=c(rgb(0.2,0.5,0.5,0.4),
          rgb(0.5,0.5,0.5,0.2)), 
  plwd=c(4,2), plty=c(1,2),
  #custom the grid
  cglcol="grey", 
  cglty=1, axislabcol="grey", 
  cglwd=0.8,
  #custom labels
  vlcex=0.8)
legend(x=-1.75, y=1.25, 
       legend = c('Cluster1'), 
       bty = "n", pch=20, col=rgb(0.2,0.5,0.5,0.4), 
       text.col = 'black', cex=1, pt.cex=3)
```


```{r spider_plot_clust2, include=TRUE, fig.height=6}
spider_plot_clust2 <- radarchart(
  bind_rows(df_cluster_summary_bind[c(1:2,4),], global_mean), 
  axistype=1, 
  #custom polygon
  pcol= c(rgb(0.8,0.2,0.5,0.4),
          rgb(0.5,0.5,0.5,0.3)), 
  pfcol= c(rgb(0.8,0.2,0.5,0.4),
           rgb(0.5,0.5,0.5,0.2)), 
  plwd=c(4,2), plty=c(1,2),
  #custom the grid
  cglcol="grey", 
  cglty=1, axislabcol="grey", 
  cglwd=0.8,
  #custom labels
  vlcex=0.8)
legend(x=-1.75, y=1.25, 
       legend = 'Cluster2', 
       bty = "n", pch=20, col=rgb(0.8,0.2,0.5,0.4), text.col = 'black', cex=1, pt.cex=3)
```



```{r spider_plot_clust3, include=TRUE, fig.height=6}
spider_plot_clust3 <- radarchart(
  bind_rows(df_cluster_summary_bind[c(1:2,5),], global_mean), 
  axistype=1, 
  #custom polygon
  pcol= c(rgb(0.7,0.5,0.1,0.4),
          rgb(0.5,0.5,0.5,0.3)),
  pfcol= c(rgb(0.7,0.5,0.1,0.4),
           rgb(0.5,0.5,0.5,0.2)), 
  plwd=c(4,2), plty=c(1,2),
  #custom the grid
  cglcol="grey", 
  cglty=1, axislabcol="grey", 
  cglwd=0.8,
  #custom labels
  vlcex=0.8)
legend(x=-1.75, y=1.25, 
       legend = 'Cluster3', 
       bty = "n", pch=20, col=rgb(0.7,0.5,0.1,0.4), text.col = 'black', cex=1, pt.cex=3)
```


# Compiling Excel of results for analysis

```{r compile_results, include=TRUE}
compile_results <- list('3clust_agnes' = cluster_agnes_groups_summary_tbl,
                        '3clust_PAM' = cluster_pam_groups_summary_tbl,
                        'membership' = joined_set,
                        'similarity' = most_similar,
                        'most_dissimilar_same_clust' = most_dissimilar_same_clust_agnes,
                        'most_dissimilar_same_clust_PAM' = most_dissimilar_same_clust_PAM,
                        'medoid_characteristics_PAM' = medoid_characteristics_PAM,
                        'medoid_characteristics_agnes' = medoid_characteristics_agnes)

write.xlsx(compile_results, file='.//output/3clust_eucl_prio_vdem.xlsx', overwrite=TRUE)
```
